---
title: "DATA 607 Extra Credit"
author: "Susanna Wong"
date: "2023-03-19"
output:
  html_document:
    theme: cerulean
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extra Credit Prompt

Working with the two JSON files available through the API at nobelprize.org, ask and answer 4 interesting questions, e.g. "Which country "lost" the most nobel laureates (who were born there but received their Nobel prize as a citizen of a different country)?"

# Retreive Data from API {.tabset}

## Prize Data

```{r }
library(jsonlite)
library(DT)

prize_url <- "http://api.nobelprize.org/v1/prize.json"
raw_prize_data <- fromJSON(prize_url)
raw_prize_data <- as.data.frame(raw_prize_data)

datatable(head(raw_prize_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

## Laureate Data

```{r }
laureate_url <- "http://api.nobelprize.org/v1/laureate.json"
laureate_raw_data <- fromJSON(laureate_url)

#laureate_raw_data 

# laureate_raw_data %>% 
#   select(id, firstname, surname, born, died, bornCountry, bornCountryCode, bornCity, diedCountry, diedCountryCode, diedCity, gender, prizes ) 


laureate_raw_data <- as.data.frame(laureate_raw_data)


datatable(head(laureate_raw_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

## Country Data

```{r }
country_url <- "http://api.nobelprize.org/v1/country.json"
raw_country_data <- fromJSON(country_url)
raw_country_data <- as.data.frame(raw_country_data)

datatable(head(raw_country_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)

```

# Tidy the Data {.tabset}

## Prize Data {.tabset}

### UnNest Dataframe Within the Laureates column

```{r, message=FALSE, warning=FALSE}
#library(tidyr)
library(tidyverse)
prize_data <- raw_prize_data %>% 
    mutate_if(is.list, map, as_data_frame) %>% 
    unnest()
# Source: https://community.rstudio.com/t/understanding-unnest-and-its-expectations/428/2

colnames(prize_data) <- c("year","category","id", "firstname","surname","motivation","share","overall motivation")

datatable(head(prize_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

### Remove Overall Motivation column

The majority of the overall motivation column does not contain any value.
This column will be remove.

```{r}
prize_data <- prize_data[,-8]

datatable(head(prize_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

## Laureate Data {.tabset}

### Unnest Dataframe Within the Laureates prize column

```{r, warning=FALSE}
laureate_data <- laureate_raw_data %>% 
    mutate_if(is.list, map, as_data_frame) %>% 
    unnest()
# laureate_data
# laureate_data %>%
#   unchop(affiliations)
# 
# laureate_data2 <- laureate_data %>%
#     mutate_if(is.list, map, as_data_frame) %>%
#     unnest()
# 
# laureate_data

colnames(laureate_data) <- c("id", "firstname","surname","born","death","born_country","born_country_code","born_city","death_country","death_country_code","death_city","gender","prize_year","prize_category","prize_share","prize_motivation","affiliations","overallMotivation")

datatable(head(laureate_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

### Split Birth and Death date columns into Year, Month, Day

```{r}
library(dplyr)
library(tidyr)
laureate_data <- laureate_data %>% 
   separate(born,c('born_year', 'born_month',"born_day"),sep="-")%>% 
   separate(death,c('death_year', 'death_month',"death_day"),sep="-")

datatable(head(laureate_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

## Country Data {.tabset}

### Rename column names

```{r }
country_data <- raw_country_data
colnames(country_data) <- c("country","country_code")

datatable(head(country_data, 50),
  plugins = "ellipsis",
  options = list(scrollX = TRUE,
    columnDefs = list(list(
      targets = "_all",
      render = JS("$.fn.dataTable.render.ellipsis(30, false )")
    ))
  )
)
```

# Question {.tabset}

## Question 1 {.tabset}

Which country has the most Nobel Prize winner by birth?

The top three countries that had the most Nobel Prize winner by birth are USA, United Kingdom, and Germany.

### Datatable

```{r}
#unique(laureate_data$born_country)

born_country_winners <- laureate_data
born_country_winners$born_country[grep("(?i)French Algeria (now Algeria)", born_country_winners$born_country)] <- "Algeria"
born_country_winners$born_country[grep("(?i)Prussia", born_country_winners$born_country)] <- "Germany"
born_country_winners$born_country[grep("(?i)Hesse-Kassel", born_country_winners$born_country)] <- "Germany"
born_country_winners$born_country[grep("(?i)Schleswig (now Germany)", born_country_winners$born_country)] <- "Germany"
born_country_winners$born_country[grep("(?i)West Germany", born_country_winners$born_country)] <- "Germany"
born_country_winners$born_country[grep("Bavaria" , born_country_winners$born_country)] <- "Germany"
born_country_winners$born_country[grep("Mecklenburg" , born_country_winners$born_country)] <- "Germany"

born_country_winners$born_country[grep("(?i)Russian Empire", born_country_winners$born_country)] <- "Netherlands"
born_country_winners$born_country[grep("Germany (now Poland)", born_country_winners$born_country)] <- "Poland"
born_country_winners$born_country[grep("Germany (now Poland)", born_country_winners$born_country)] <- "Poland"
born_country_winners$born_country[grep("(?i)Austria-Hungary", born_country_winners$born_country)] <- "Poland"
born_country_winners$born_country[grep("(?i)German-occupied Poland", born_country_winners$born_country)] <- "Poland"
born_country_winners$born_country[grep("(?i)Free City of Danzig", born_country_winners$born_country)] <- "Poland"
born_country_winners$born_country[grep("(?i)the Netherlands", born_country_winners$born_country)] <- "Poland"

#unique(born_country_winners$born_country)

born_country_winners <- born_country_winners  %>% 
  group_by(born_country)%>% 
   count(born_country)%>%
   mutate(percentage = (n / nrow(laureate_data))*100)

born_country_winners$percentage <- as.numeric(format(round(born_country_winners$percentage,2),nsmall =2))

datatable(born_country_winners)
```

### Visualization

```{r, warning=FALSE}
born_country_winners <- distinct(born_country_winners) %>%
  filter(percentage > 2 )

born_country_winners %>% 
    arrange(desc(n)) %>%
    ggplot(., aes(x = reorder(born_country, n), y = n)) +
  geom_bar(stat = 'identity', skill= "blue" , fill = 'lightblue') +
  coord_flip() + ggtitle("Top 10 General Skills") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Nobel Prize by Country") +
  xlab("Count")

```

## Question 2 {.tabset}

There are 6 prize category: "physics", "chemistry", "peace","medicine", "literature", and "economics".
A female, male, or organization can receive the Nobel Prize.
What is the distribution of gender receiving a Nobel prize for each category?

For each category, male has the highest percentage of receiving the Nobel Prize.
Only organizations have receive a Nobel Prize in the the peace category.

### Datatable

```{r}
unique(laureate_data$gender)
unique(laureate_data$prize_category)
novel_prize_category <- laureate_data %>% 
  group_by(prize_category)%>% 
   count(gender)%>%
   mutate(percentage = (n / nrow(laureate_data))*100)

novel_prize_category$percentage <- as.numeric(format(round(novel_prize_category$percentage,2),nsmall =2))

datatable(novel_prize_category)
```

### Visualization

```{r}
ggplot(novel_prize_category,                                      
       aes(x = prize_category,
           y = n,
           fill = gender)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  ggtitle("Nobel Prize by Category") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Prize Category") +
  xlab("Count")
```

## Question 3

Although the Novel Prize has been received by predominately male, has the number of prizes received by the other 'genders' increase a lot over time?

The number of females receiving a Novel Prize did increase over time after 2000.

```{r}
gender_over_time <- laureate_data[c("prize_year","gender")] %>%
  group_by(gender) %>%
  mutate(count=row_number())
  # mutate(counter = seq_along(gender)) %>%
  # ungroup()

gender_over_time$prize_year <- as.Date(as.character(gender_over_time$prize_year),
                        format = "%Y")

ggplot(gender_over_time, aes(x = prize_year, y = count, colour = gender)) +
  geom_smooth() +scale_x_date(date_labels = "%Y ")
```

## Question 4 {.tabset}

What is the percentage of laureate shared their prize?

About 64% of the laureate shared their Nobel Prize. So far, the Nobel Prize can be shared up to 4 laureates. 
### Datatable

```{r}
unique(laureate_data$prize_share)

novel_prize_share <- laureate_data %>% 
  group_by(prize_share)%>% 
   count(prize_share)%>%
   mutate(percentage = (n / nrow(laureate_data))*100)

novel_prize_share$percentage <- as.numeric(format(round(novel_prize_share$percentage,2),nsmall =2))

datatable(novel_prize_share)

```

### Visualization

```{r}
ggplot(novel_prize_share,                                      
       aes(x = prize_share,
           y = percentage,)) +
  geom_bar(stat = 'identity', skill= "blue" , fill = 'lightblue') +
  ggtitle("Nobel Prize Share") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Count") +
  xlab("Prize Share")


```

## Question 5

Extension to question 4:
Which category did not have a high prize share? 

The majority of laureate in literature did not share their Nobel Prize. 
### Datatable

```{r}
unique(laureate_data$prize_share)

novel_prize_share_category <- laureate_data %>% 
  group_by(prize_category,prize_share)%>% 
   count(prize_share)%>%
   mutate(percentage = (n / nrow(laureate_data))*100)

novel_prize_share_category$percentage <- as.numeric(format(round(novel_prize_share_category$percentage,2),nsmall =2))

datatable(novel_prize_share_category)

```

### Visualization

```{r}
ggplot(novel_prize_share_category,                                      
       aes(x = prize_category,
           y = n,
           fill = prize_share)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  ggtitle("Nobel Prize by Category") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Count") +
  xlab("Prize Category")
```

## Question 6
What percentage of the laureate are still alive?

33% of laureate are still alive.
```{r}
alive <- laureate_data %>%
  filter(death_year=="0000") %>%
  count()

alive_percentage = (alive$n / nrow(laureate_data))*100
alive_percentage


# Conclusion

1.  Which country has the most Nobel Prize winner by birth?

The top three countries that had the most Nobel Prize winner by birth are USA, United Kingdom, and Germany.

2.  There are 6 prize category: "physics", "chemistry", "peace","medicine", "literature", and "economics". A female, male, or organization can receive the Nobel Prize. What is the distribution of gender receiving a Nobel prize for each category?

For each category, male has the highest percentage of receiving the Nobel Prize.
Only organizations have receive a Nobel Prize in the the peace category.

3.  Although the Novel Prize has been received by predominately male, has the number of prizes received by the other 'genders' increase a lot over time?

The number of females receiving a Novel Prize did increase over time after 2000.

4. What is the percentage of laureate shared their prize?

About 64% of the laureate shared their Nobel Prize. So far, the Nobel Prize can be shared up to 4 laureates. 

5. Extension to question 4: Which category did not have a high prize share? 

The majority of laureate in literature did not share their Nobel Prize. 

6. What percentage of the laureate are still alive?

33% of laureate are still alive.
